---
title: "VOCs Vendy"
author: "Anna Mrazova"
date: "2025-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# VOC analyses

## Used libraries

```{r libraries}
library(Matrix)
library(lme4)
library(dplyr)
library(tidyverse)
library(car)
library(glmmTMB)
library(emmeans)
library(MuMIn)
library(DHARMa)
library(vegan)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(dplyr)
library(tidyr)
library(tibble)
```

## 1 STROMY BEZ MeJA
## A) mnozství VOC ~ druh

### Data filtering
```{r data 1A}
setwd("C:/Users/annam/OneDrive - Biologické centrum AV ČR, v. v. i/Anawork/Students/Vendula Bilkova/Data/voc")
data1A <- read.csv2("RStudio data 1.csv", dec = ",", header = TRUE, stringsAsFactors = FALSE)
data1A <- subset(data1A, grepl("kontr_pid|kontr_ostruh", Treatment, ignore.case = TRUE))
data1A <- subset(data1A, Compound == "Total_Standard")
data1A$Treatment <- as.factor(data1A$Treatment)
data1A$Value <- as.numeric(data1A$Value)
```

### Perform appropriate test
#### Based on data diagnosis tests:
#### **1)** check for normality and equal variaces hold, we we go for a simple t-test
#### **2)** If normality is violated, go for non-parametric wilcoxon test
```{r test 1A}
# Check normality assumption per group
shapiro_test_ostr <- shapiro.test(data1A$Value[data1A$Treatment == "kontr_ostruh"])
shapiro_test_pid <- shapiro.test(data1A$Value[data1A$Treatment == "kontr_pid"])

if (shapiro_test_ostr$p.value > 0.05 & shapiro_test_pid$p.value > 0.05) {
  if (levene_test$`Pr(>F)`[1] > 0.05) {
       test_result <- t.test(Value ~ Treatment, data = data1A, var.equal = TRUE)
  }
} else {
    test_result <- wilcox.test(Value ~ Treatment, data = data1A)
}
print(test_result)
```

### Visualisation 
```{r visualisation 1A}
ggplot(data1A, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Comparison of VOC abundance between treatments",
       y = "Total VOC abundance",
       x = "Treatment")
```

### Results t-test
#### wilcoxon test for non normally distributed data - **treatment has no effect**.

## B) an overall VOC variation between treatments - PCA
### Data manipulation
```{r data 1B}
data1B <- read.csv2("RStudio data 1.csv", dec = ",", header = TRUE, stringsAsFactors = FALSE)
data1B <- subset(data1B, grepl("kontr_pid|kontr_ostruh", Treatment, ignore.case = TRUE))
data1B <- data1B %>% filter(Compound != "Total_Standard")

# log transformation for values
data1B$log_Value <- log(data1B$Value)

# Prepare VOC data in wide format
VOC_matrix <- data1B %>%
  select(ID, Compound, log_Value) %>%
  pivot_wider(names_from = Compound, values_from = log_Value) %>%
  left_join(data1B %>% select(ID, Treatment) %>% distinct(), by = "ID") %>%
  column_to_rownames(var = "ID")

# Convert Treatment to a factor
VOC_matrix$Treatment <- as.factor(VOC_matrix$Treatment)
```

### PCA 1
```{r PCA1}
# Run PCA excluding the Treatment column
pca_result <- PCA(VOC_matrix %>% select(-Treatment), scale.unit = TRUE, graph = FALSE)

# Visualize PCA with treatment groups
palette_colors <- c("red", "blue")
fviz_pca_ind(pca_result, 
             label = "none", 
             habillage = VOC_matrix$Treatment,
             addEllipses = TRUE,  
             ellipse.level = 0.95,
             palette = palette_colors)
```

### Results PCA
#### **Separation Between Treatments:**
#### The two groups (red: kontr_ostruh, blue: kontr_pid) show partial separation. There is some overlap, but kontr_pid (blue) seems to be more dispersed, while kontr_ostruh (red) is more concentrated.

#### **Main Variation Explained by Dim1 (41.2%):**
#### The first principal component (Dim1) explains 41.2% of the variation in VOC profiles. This suggests that a large proportion of VOC variation is associated with treatment differences. The separation along Dim1 suggests that kontr_ostruh and kontr_pid may emit VOCs at different intensities or compositions.

#### **Dim2 (15.4%) Shows Additional Variation:**
#### There is some vertical spread, but it’s less structured compared to Dim1. This might reflect individual variation rather than a treatment effect.

#### **Ellipses Indicate Group Variability:**
#### The blue ellipse (kontr_pid) is larger, suggesting higher within-group variance. The red ellipse (kontr_ostruh) is more compact, meaning these samples have more similar VOC profiles.

## C) Effect of treatment (species feeding) on individual compounds

### Fitting model and variable selection
``` {r models 1C}
model1C <- glmer(Value ~ Treatment * Compound + (1|ID), data = data1B, family = Gamma(link = "log"))

model1C <- glmmTMB(Value ~ Treatment * Compound + (1|ID), data = data1B, family = Gamma(link = "log"))

reduced1 <- update(model1C, . ~ . - Compound)                 
reduced2 <- update(model1C, . ~ . - Treatment)              
reduced3 <- update(model1C, . ~ . - Compound:Treatment)
Null <- glmmTMB(Value ~ 1 + (1|ID), data = data1B, family = Gamma(link = "log"))

# Compare models using AICc
models_list <- list(model1C, reduced1, reduced2, reduced3, Null)
model_selection <- model.sel(models_list, rank = AICc)
print(model_selection)

# Diagnosis check
simulationOutput <- simulateResiduals(fittedModel = model1C, n = 1000)
plot(simulationOutput)
testDispersion(simulationOutput)
hist(data1B$Value, breaks = 30, main = "Histogram of VOC abundance", xlab = "Value")

# -> log transformarion
data1B$log_Value <- log(data1B$Value)

model_logT <- glmmTMB(log_Value ~ Treatment * Compound + (1|ID), 
                      data = data1B, family = gaussian)
simulationOutput_logT <- simulateResiduals(fittedModel = model_logT, n = 1000)
plot(simulationOutput_logT)

summary(model_logT)

emmeans(model_logT, pairwise ~ Treatment, by = "Compound", type = "response")
```

### Results

#### A post-hoc analysis was conducted using estimated marginal means (EMMs) to compare VOC levels between treatments for each compound. Some key findings:

#### **Significant Differences:**
#### **Methyl salicylate:** Significantly higher in kontr_ostruh than in kontr_pid (p=0.0006p=0.0006).
#### ***Unknown 19:** lower in kontr_pid (p=0.0497p=0.0497), suggesting a species-specific reduction.

#### **Non-Significant Differences (but):**
#### Many other compounds showed minor differences but did not reach statistical significance. Some unknown compounds (Unknown 10, 11, 14, 21) showed a tendency towards treatment-related variation, but their confidence intervals overlapped.

### Final Interpretation:

#### The overall effect of trvidual compounds is weak, but specific VOCs, particularly Methyl salicylate and Unknown 19, appear to differ significantly.
#### The PCA earlier suggested that VOC profiles are structured by treatment, but when zooming in on individual compounds, many do not differ significantly.
#### The treatment effect likely depends on the compound, suggesting that some VOCs are more sensitive to herbivory by specific species.
#### Implications: These results indicate that herbivore species identity may influence VOC emissions, but only for select compounds.

## 2) POROVNANI VOC u MeJA vs zrano housenkou
# A) mnozstvi VOC ~ druh housenky * treatment + (1|strom)

### Data filtering
```{r data 2A}
data2A <- read.csv2("RStudio data 1.csv", dec = ",", header = TRUE, stringsAsFactors = FALSE)
data2A <- subset(data2A, Compound == "Total_Standard")
data2A <- data2A %>% filter(Treatment %in% c("meja_ostruh", "meja_pid"))
data2A$Treatment <- as.factor(data2A$Treatment)
data2A$Value <- as.numeric(data2A$Value)
```

### Perform appropriate test
#### Based on data diagnosis tests:
#### **1)** check for normality and equal variaces hold, we we go for a simple t-test
#### **2)** If normality is violated, go for non-parametric wilcoxon test
```{r test 2A}
# Check normality assumption per group
shapiro_test_ostr <- shapiro.test(data2A$Value[data2A$Treatment == "meja_ostruh"])
shapiro_test_pid <- shapiro.test(data2A$Value[data2A$Treatment == "meja_pid"])

if (shapiro_test_ostr$p.value > 0.05 & shapiro_test_pid$p.value > 0.05) {
  if (levene_test$`Pr(>F)`[1] > 0.05) {
    test_result <- t.test(Value ~ Treatment, data = data2A, var.equal = TRUE)
  }
} else {
  test_result <- wilcox.test(Value ~ Treatment, data = data2A)
}

print(test_result)
```

### Visualisation 
```{r visualisation 2A}
ggplot(data2A, aes(x = Treatment, y = Value, fill = Treatment)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Comparison of VOC abundance between treatments",
       y = "Total VOC abundance",
       x = "Treatment")
```

### Results t-test 2A
#### wilcoxon test for non normally distributed data - **treatment has a significant effect**.

## 2 B) an overall VOC variation between treatments - PCA
### Data manipulation
```{r data 2B}
data2B <- read.csv2("RStudio data 1.csv", dec = ",", header = TRUE, stringsAsFactors = FALSE)
data2B <- data2B %>% filter(Treatment %in% c("meja_ostruh", "meja_pid"))
data2B <- data2B %>% 
  filter(Compound != "Total_Standard")

data2B$log_Value <- log(data2B$Value)

VOC_matrix <- data2B %>%
  select(ID, Compound, log_Value) %>%
  pivot_wider(names_from = Compound, values_from = log_Value) %>%
  left_join(data2B %>% select(ID, Treatment) %>% distinct(), by = "ID") %>%
  column_to_rownames(var = "ID")

VOC_matrix$Treatment <- as.factor(VOC_matrix$Treatment)
```

### PCA 2
```{r PCA2}
pca_result <- PCA(VOC_matrix %>% select(-Treatment), scale.unit = TRUE, graph = FALSE)

palette_colors <- c("red", "blue")

fviz_pca_ind(pca_result, 
             label = "none", 
             habillage = VOC_matrix$Treatment,
             addEllipses = TRUE,  
             ellipse.level = 0.95,
             palette = palette_colors)
```

### Results PCA

#### **Separation Between Treatments:**
#### The first principal component (Dim1, 47.6%) captures nearly half of the variation in the dataset.Meja_ostruh (red) and meja_pid (blue) are separated along this axis, indicating that VOC profiles are systematically different between treatments.Meja_pid shows less variation and is more tightly clustered, suggesting more consistent VOC production. Meja_ostruh has a wider spread, indicating greater variability in VOC composition.

#### **Ellipses Represent Confidence Intervals (95% CI):**
#### The red ellipse (for meja_ostruh) is larger, suggesting more variation in VOC profiles. The blue ellipse (meja_pid) is tighter and more compact, meaning lower variation in VOC responses. The overlap is small, reinforcing that meja_ostruh and meja_pid have distinct VOC profiles.

#### **Biological Interpretation:**
#### meja_ostruh appears to induce a stronger and more variable VOC response, which might be linked to plant defense activation. Meja_pid, in contrast, seems to reduce VOC diversity or stabilize the response, potentially suppressing certain compounds.

## C) Effect of treatment (species feeding) on individual compounds when MeJA treated

### Fitting model and variable selection
``` {r models 2C}

model2B <- glmer(Value ~ Treatment * Compound + (1|ID), data = data2B, family = Gamma(link = "log"))
#convergence issue ->

model2B <- glmmTMB(Value ~ Treatment * Compound + (1|ID), data = data2B, family = Gamma(link = "log"))

reduced1 <- update(model2B, . ~ . - Compound)                 
reduced2 <- update(model2B, . ~ . - Treatment)              
reduced3 <- update(model2B, . ~ . - Compound:Treatment)
Null <- glmmTMB(Value ~ 1 + (1|ID),  
                data = data2B, family = Gamma(link = "log"))

# Compare models using AICc
models_list <- list(model2B, reduced1, reduced2, reduced3, Null)
model_selection <- model.sel(models_list, rank = AICc)
print(model_selection)

simulationOutput <- simulateResiduals(fittedModel = model2B, n = 1000)
plot(simulationOutput)
testDispersion(simulationOutput)

hist(data2B$Value, breaks = 30, main = "Histogram of VOC Abundance", xlab = "Value")

# data are scewed -> use the log transformation

model_logT <- glmmTMB(log_Value ~ Treatment * Compound + (1|ID), 
                      data = data2B, family = gaussian)
simulationOutput_logT <- simulateResiduals(fittedModel = model_logT, n = 1000)
plot(simulationOutput_logT)

summary(model_logT)

emmeans(model_logT, pairwise ~ Treatment, by = "Compound", type = "response")
```

### Results

#### **1. Key Ecological Question:**
#### Which VOC compounds show a significantly different response depending on the treatment (meja_pid vs. meja_ostruh), and what does this mean ecologically?
  
#### **2. How to Identify Significant Interactions**
#### We focus on interaction terms in the model: Treatment×Compound
#### This tells us whether the effect of meja_pid depends on the specific compound.

#### **Significant interaction term:**
#### Treatmentmeja_pid:CompoundMethyl_salicilate_Standard (p = 0.033)
#### Interpretation: The effect of meja_pid differs significantly for Methyl salicylate compared to meja_ostruh.

#### **Near-significant interactions (0.05 < p < 0.1):**
#### Treatmentmeja_pid:CompoundUnknown_13_Standard (p = 0.072)
#### Treatmentmeja_pid:CompoundUnknown_6_Standard (p = 0.068)
#### These might still be biologically meaningful, even if they don’t reach strict statistical significance.

#### **3. Breakdown of pairwise comparisons (emmeans)**
#### Which compounds show a strong difference between meja_pid and meja_ostruh?

#### **Strongly suppressed in meja_pid:**

#### Unknown_1_Standard (p = 0.0002)
#### meja_ostruh: 12.38, meja_pid: 10.57

#### Unknown_14_Standard (p = 0.0055)
#### meja_ostruh: 14.10, meja_pid: 12.74

#### Unknown_21_Standard (p = 0.0001)
#### meja_ostruh: 14.23, meja_pid: 12.29

#### 2-Decen-1-ol_Standard (p = 0.036)
#### meja_ostruh: 16.67, meja_pid: 15.65

#### Butan-1-ol_Standard (p = 0.0078)
#### meja_ostruh: 16.18, meja_pid: 14.88

#### **Near-significant**
  
#### Unknown_13_Standard (p = 0.072)
#### Unknown_6_Standard (p = 0.068)
#### Unknown_18_Standard (p = 0.074)

#### **Takeaway**: Certain unknown compounds and well-known defensive VOCs (e.g., Methyl salicylate, Butan-1-ol) are strongly affected by the interaction between treatment and compound type.

#### **4. Ecological Interpretation**
#### 1) What does meja_pid do to VOCs?
#### Many defensive compounds (Methyl salicylate, Butan-1-ol) and unknowns are significantly reduced under meja_pid.
#### This could suggest that **meja_pid suppresses VOC emissions involved in plant defense.**

#### 2) What about meja_ostruh?
#### meja_ostruh consistently shows higher levels of VOCs, suggesting it maintains or increases VOC emissions.
